A Large-scale Study of Polymorphism in R

### 2019-03-28 paper/main.tex
small code updates to deal with failed traces, and light reformatting of paper, writing a bit about function level polymorphism

@@

The first question we will address is: {\it how polymorphic is R?}  To
answer that, we will first look at the number of monomorphic arguments
across all of the packages we analyzed.  The data can be found in
Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}
\AT{TODO: put all counts in one table?}

@@

Another option is a date.
In R, dates are often represented as either strings or numbers, and many packages make use of dates.

\AT{We might want to discuss some of the most polymorphic functions in the language.}
\subsection{Most Polymorphic Functions}

@@

%
%
%
%
\subsection{Types in R}

\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}

%
%
\subsubsection{What are possible type systems?}

Great question.
Some things:

\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs. a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.

\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.

\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
This gives us ... .

Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:

\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.

\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.

\item {\bf rolling NA into other types}: \AT{i should try this}
similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.

\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.

%
%
%
%
\subsection{Usage Patterns}

In this section, we will discuss function usage patterns which arose in our analysis.
We will start by looking at the morphicity of functions.

\subsubsection{Function Argument and Return Morphicity}

First and foremost, we would like to know how often R programmers create polymorphic functions.
Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.

Here, we see that the vast majority of function arguments are indeed monomorphic.
Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
That said, types alone don't always paint the whole picture:
recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
So in which circumstances \textit{do} types paint the whole picture?

The second data point in \textbf{TODO FIGURE} indicates that a
\isit{majority} of function arguments are monomorphic in type \textit{and}
have no attributes.  These represent arguments which are truly trivial to
annotate, as the {\tt typeof} an argument perfectly describes the usage of
that argument.  That said, some attributes arise naturally in R: For
instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
  y=0)}) appear as an attribute on the value.  In
Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
that a nontrivial amount of functions are monomorphic in type with said
natural attribute patterns.

Another facet of type information in R is in the {\tt class} attribute.
Recall that values have a \textit{class} in addition to a type: For example,
a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
\todo{Plug classes earlier} R has a number of built-in classes, such as
\todo{X, Y, and Z}, but users are free to redefine the class of any value at
runtime, and easily define new classes.  The next data point in
\todo{Figure} shows that user-defined classes don't appear altogether often,
though they do indeed feature.  \AT{In a following section, we will discuss
  how the usage of these classes manifests itself.}

\AT{Separate section for functions? Right now, this is at argument granularity.}

\subsubsection{Type Signatures}

In the last section, we presented a high-level overview of the morphicity ... .

\subsubsection{Attribute Signatures}

Recall that attributes are a way for programmers to store metadata on values in R.
What are the common attribute patterns?
And how often is the attribute pattern polymorphic?

\subsubsection{Takeaways}

\begin{itemize}
    \item the vast majority of arguments are monomorphic in type;
    \item of those, over 60\% have no attribute information;
    \item of the 40\% with attribute information, roughly 1/2 have fairly
      simple attributes corresponding to base R constructs (named lists and
      vectors, matrices, and data frames);
    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
\end{itemize}

In short, it looks like R (package) programmers are reasonable.  I'd
conjecture that a lot of the polymorphism (e.g., double and list) is coming
from how easy it is to use either type in a given situation (e.g.,
converting from vector of doubles to list or vice versa is simple).

%
%
%
%
%
%
\section{Synthesis}

In this section, we will discuss the conclusions that we draw from our data.

%
%
\subsection{Suggested Annotations}

\AT{How can we capture these patterns with annotations?}

Some possible annotations:

\begin{itemize}
    \item \textit{real}: for \textit{double} and \textit{integer} values
    \item \textit{function}: for \textit{closure}, \textit{special}, and \textit{builtin} values
    \item \textit{vector}: to indicate that something should be vectorized
    \item \textit{scalar}: to indicate that something should \textbf{not} be vectorized
    \item \textit{index}: for \textit{real} and \textit{character} values
\end{itemize}

%
%
\subsubsection{Struct-Like Attribute Declarations}

\AT{What's a convenient way to annotate attributes?  Should investigate how
  often attributes are consistent.}

%
\subsubsection{Coverage}

@@

>>>>

\AT{Do we want to talk about how sometimes we can't eval an argument, even if a function executes successfully?
Like pipe from magrittr.}

@@

Broadly, the first question we will address is: {\it how polymorphic is R?} 
We will answer this in two parts, and first we will look at the number of monomorphic {\it arguments} across all of the packages we analyzed.

%
%
%
%
\subsection{Argument Polymorphism}

In this section, we will explore the polymorphicity of function argument positions and return values.
Exploring argument polymorphism will help to motivate a design for whole function signatures.

To begin, let's look at the {\it amount} of polymorphism by way of {\it number of polymorphic arguments}.
The data can be found in Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}

@@

Another option is a date.
In R, dates are often represented as either strings or numbers, and many packages make use of dates.
Dates (sometimes) manifest themselves in the {\tt class} of a value, for examples refer to Figure~\ref{fig:date}.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
class(date()) # => character
class(Sys.Date()) # => Date
class(as.POSIXlt(Sys.Date())) # => POSIXlt, POSIXt
as.integer(as.Date(1, origin = "1900-01-01"))
\end{lstlisting}}\caption{Dates in R.}\label{fig:date}\end{figure}

This figure reveals an unfortunate truth that date ``data type'' in R is not standardized.
That said, translating between the data types is straigthforward.

%
%
%
%
\AT{We might want to discuss some of the most polymorphic functions in the language.}
\subsection{Most Polymorphic Functions}

@@

...

%
%
%
%
\subsection{Function Polymorphism}

Now, we will turn our attention to whole function polymorphism.

This is different because we can now look at argument polymorphism across several arguments, like we did when thinking about character/real polymorphism.

We're interested to see what kinds of ``meta-annotations'' we can make at the function level.
For instance, should we link argument types with return types?
Consider Figure~\ref{fig:funanno}.
\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
iden <- function(x) x;

# observed calls:
f(2); # x: double -> double
f(list(1, 2, 3)) # x: list<double> -> list<double>

# possible annotations:
# [1] x: {double, list<double>} -> {double, list<double>}
# [2] x: {double -> double, x: list<double> -> list<double>}
\end{lstlisting}}\caption{Possible function annotations.}\label{fig:funanno}\end{figure}
The first annotation option is

%
%
%
%
\subsection{Coverage}

@@

%%
%%
%%
%%
%\subsection{Types in R}
%
%\AT{I'll pull from this section when appropriate, as the discussion of the type systems becomes relevant.}
%
%%
%%
%\subsubsection{What are possible type systems?}
%
%Great question.
%Some things:
%
%\begin{itemize}
%
%\item {\it The Obvious Type System (OTS)}: this type system corresponds directly with the type information that one can fetch using the {\tt typeof}, {\tt attributes}, and {\tt class} functions.
%In this system, there is no distinction between vectors and scalars, and lists of differently-typed elements (e.g., a list of integers vs. a list of strings), and there is a distinction between integers and doubles, and the type of NULL and all other types.
%This type system isn't particularly informative to a programmer, though it does reflect the treatment of values in the R runtime, for the {\tt typeof} a value is the type that the value has according to the R internals.
%
%\item {\it The Fine-Grained Type System (FGTS)}: this type systems includes a number of relevant additions.
%Chiefly, it distinguishes between vectors and scalars, as it is conceivable that boxing scalars as single-element vectors might be overkill.
%In addition, list types are parameterized over the types of their elements (e.g., list<integer> vs. list<character>).
%Finally, NAs ``in the wild'' are given the type {\tt raw\_NA}, as {\tt typeof(NA) = logical} in R.
%
%\item {\it The Null-and-NA-Free Type System (NNTS)}: most languages consider NULLs to inhabit all types, and R does so for NAs \AT{elaborate}.
%This gives us ... .
%
%\end{itemize}
%
%Of course, one could conceive of a more fine-grained type system than FGTS, but it's unclear that we would gain much in doing so.
%Indeed, there is a space to explore in slightly less fine-grained type systems, which make modifications including:
%
%\begin{itemize}
%
%\item {\bf collapsing vectors and scalars}: this yields a type system slightly closer to OTS.
%
%\item {\bf rolling NULL into other types}: in most languages, NULL is an inhabitant of all types, and this adjustment seems interesting.
%Note that this goes beyond even OTS, as in R {\tt typeof(NULL) = NULL}.
%
%\item {\bf rolling NA into other types}: \AT{i should try this}
%similarly, NA is an inhabitant of all types, though NA standing alone has type {\tt logical}.
%FGTS distinguishes such NAs with the {\tt raw\_NA} type, and if instead we consider NA to inhabit all types we may \AT{win}.
%
%\item {\bf combining integers and doubles}: as in most languages, doubles and ints can stand-in for each other in nearly every scenario.
%Even if not, it is trivial to convert from one to the other, particularly converting ints to doubles.
%Aside, in R a double can be used to index a list, and R will floor it when performing the lookup.
%
%\end{itemize}
%\subsection{Usage Patterns}
%
%In this section, we will discuss function usage patterns which arose in our analysis.
%We will start by looking at the morphicity of functions.
%
%\subsubsection{Function Argument and Return Morphicity}
%
%First and foremost, we would like to know how often R programmers create polymorphic functions.
%Recall that we define a polymorphic function to be a function with at least one polymorphic argument, or a polymorphic return.
%We will turn our attention now to the data in \todo{Figure}, and go through each entry in the table.
%
%Here, we see that the vast majority of function arguments are indeed monomorphic.
%Monomorphic arguments are easy to annotate, as the {\tt typeof} the arguments is exactly the most precise annotation we could give, at least in terms of \textit{type}.
%That said, types alone don't always paint the whole picture:
%recall that \textit{attributes} are an R language feature which allows programmers to stick metadata onto values.
%So in which circumstances \textit{do} types paint the whole picture?
%
%The second data point in \textbf{TODO FIGURE} indicates that a
%\isit{majority} of function arguments are monomorphic in type \textit{and}
%have no attributes.  These represent arguments which are truly trivial to
%annotate, as the {\tt typeof} an argument perfectly describes the usage of
%that argument.  That said, some attributes arise naturally in R: For
%instance, names in a named list (e.g. {\tt x} and {\tt y} in {\tt list(x=0,
%  y=0)}) appear as an attribute on the value.  In
%Section~\ref{sec:method:attributes}, we outlined these naturally-occurring
%attribute patterns, and the third data point in \textbf{TODO FIGURE} shows
%that a nontrivial amount of functions are monomorphic in type with said
%natural attribute patterns.
%
%Another facet of type information in R is in the {\tt class} attribute.
%Recall that values have a \textit{class} in addition to a type: For example,
%a {\tt data.frame} has type {\tt list} and class {\tt data.frame}.
%\todo{Plug classes earlier} R has a number of built-in classes, such as
%\todo{X, Y, and Z}, but users are free to redefine the class of any value at
%runtime, and easily define new classes.  The next data point in
%\todo{Figure} shows that user-defined classes don't appear altogether often,
%though they do indeed feature.  \AT{In a following section, we will discuss
%  how the usage of these classes manifests itself.}
%
%\AT{Separate section for functions? Right now, this is at argument granularity.}
%
%\subsubsection{Type Signatures}
%
%In the last section, we presented a high-level overview of the morphicity ... .
%
%\subsubsection{Attribute Signatures}
%
%Recall that attributes are a way for programmers to store metadata on values in R.
%What are the common attribute patterns?
%And how often is the attribute pattern polymorphic?
%
%\subsubsection{Takeaways}
%
%\begin{itemize}
%    \item the vast majority of arguments are monomorphic in type;
%    \item of those, over 60\% have no attribute information;
%    \item of the 40\% with attribute information, roughly 1/2 have fairly
%      simple attributes corresponding to base R constructs (named lists and
%      vectors, matrices, and data frames);
%    \item now, of those arguments which are polymorphic, a sizable chunk (well over half) have defensible signatures (e.g., double and character for named list indexing, double and integer for obvious reasons, etc.).
%\end{itemize}
%
%In short, it looks like R (package) programmers are reasonable.  I'd
%conjecture that a lot of the polymorphism (e.g., double and list) is coming
%from how easy it is to use either type in a given situation (e.g.,
%converting from vector of doubles to list or vice versa is simple).
